---
title: "Replication: Diegert, Masten and Poirier (2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replication: Diegert, Masten and Poirier (2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
options(width = 100)
```

This vignette reproduces every empirical table and figure of
**Diegert, Masten and Poirier (2026),** [*Assessing Omitted Variable Bias
when the Controls are Endogenous*][dmp], using the bundled `bfg2020`
data. We label each chunk with the corresponding table/figure number in
the paper.

[dmp]: https://arxiv.org/abs/2206.02303

```{r}
library(regsensitivity)
library(ggplot2)
data(bfg2020)
bfg2020$statea <- factor(bfg2020$statea)

w1 <- c("log_area_2010", "lat", "lon", "temp_mean", "rain_mean",
        "elev_mean", "d_coa", "d_riv", "d_lak", "ave_gyi")
labels <- c(
    log_area_2010 = "Land area",
    lat           = "Centroid Latitude",
    lon           = "Centroid Longitude",
    temp_mean     = "Average temperature",
    rain_mean     = "Average rainfall",
    elev_mean     = "Elevation",
    d_coa         = "Distance from centroid to the coast",
    d_riv         = "Distance from centroid to rivers",
    d_lak         = "Distance from centroid to lakes",
    ave_gyi       = "Average potential agricultural yield"
)
form <- avgrep2000to2016 ~ tye_tfe890_500kNI_100_l6 +
    log_area_2010 + lat + lon + temp_mean + rain_mean + elev_mean +
    d_coa + d_riv + d_lak + ave_gyi + statea
```

## Table 1, Panel C, column (5): breakdown points

For the Republican vote share outcome the paper reports:

* $\bar r_X^{bp} = 0.804$ -- the breakdown rxbar at $\bar c = 1$.
* $\bar r^{bp} = 0.96$    -- the breakdown rxbar under
  $\bar r_Y = \bar r_X$ (common maximal impact).

```{r}
bp_conservative <- regsen_bounds(form, bfg2020, compare = w1,
                                  cbar = 1)
bp_relaxed <- regsen_bounds(form, bfg2020, compare = w1,
                             cbar = 1,
                             rybar_expr = function(rx) rx)
```

Side by side against the published values. `published` holds the figures
as printed in the paper, so the comparison is checked on every build
rather than asserted in prose:

```{r bp-comparison}
# `digits` is the precision the paper actually prints each value to, given
# explicitly rather than inferred. Deriving it from format() is wrong:
# format() pads a vector to a common width, so format(c(0.804, 0.96)) is
# c("0.804", "0.960") and 0.96 would be held to a precision it was never
# printed at.
compare_published <- function(quantity, published, computed, digits) {
    out <- data.frame(
        quantity  = quantity,
        published = published,
        computed  = round(computed, 4),
        stringsAsFactors = FALSE
    )
    out$difference <- round(out$computed - out$published, 4)
    out$agrees <- abs(out$difference) <= 0.5 * 10^(-digits)
    out
}

cmp <- compare_published(
    quantity  = c("rxbar^bp at cbar = 1",
                  "r^bp at cbar = 1, rybar = rxbar"),
    published = c(0.804, 0.96),
    digits    = c(3, 2),          # as printed in the paper
    computed  = c(bp_conservative$breakdown, bp_relaxed$breakdown)
)
knitr::kable(cmp, caption = "Breakdown points: this package vs DMP (2026).")
```

```{r bp-assert, echo = FALSE}
stopifnot(all(cmp$agrees))
```

Both agree to the precision the paper prints. The `stopifnot()` above is
deliberate: if a future change moved either number, this vignette would
stop building rather than quietly publishing a wrong replication.

## Table 3: correlations between observed covariates

```{r}
tbl3 <- calibrate_partial_r2(form, bfg2020, compare = w1)
tbl3$variable <- labels[tbl3$variable]
print(tbl3, row.names = FALSE)
```

Matches the paper to three decimal places.

## Table 4, column (5): rho_k calibration

```{r}
tbl4 <- calibrate_rho(form, bfg2020, compare = w1)
tbl4$variable <- labels[tbl4$variable]
print(tbl4, row.names = FALSE)
```

Each value matches the paper's Table 4 to one decimal place.

## Figure 1: Sensitivity analysis for Republican Vote Share

### Left panel: bounds on $\beta_{long}$ as a function of $\bar r_X$, $\bar c = 1$

The paper's left panel carries three things: solid bounds with
$\bar r_Y = \infty$, **dashed** bounds under the common-maximal-impact
restriction $\bar r_Y = \bar r_X$, and tick marks on the horizontal axis
showing the Table 4 calibration values. All three are reproduced below.

```{r fig1-left, fig.height = 4.5}
rx_grid <- seq(0, 1.4, length.out = 300)
solid  <- regsen_bounds(form, bfg2020, compare = w1, cbar = 1,
                         rxbar = rx_grid)
dashed <- regsen_bounds(form, bfg2020, compare = w1, cbar = 1,
                         rxbar = rx_grid,
                         rybar_expr = function(rx) rx)

sd <- as.data.frame(solid);  sd$spec <- "rybar = Inf"
dd <- as.data.frame(dashed); dd$spec <- "rybar = rxbar"
df <- rbind(sd[, c("rxbar", "bmin", "bmax", "spec")],
            dd[, c("rxbar", "bmin", "bmax", "spec")])

# Past the asymptote the bounds are infinite. coord_cartesian() squishes
# infinite values onto the panel edge rather than dropping them, which
# paints a line along the top; the large finite values just below the
# asymptote already show the divergence, so blank the infinities.
df$bmin[!is.finite(df$bmin)] <- NA
df$bmax[!is.finite(df$bmax)] <- NA

ticks <- data.frame(x = calibrate_rho(form, bfg2020, compare = w1)$rho / 100)

ggplot(df, aes(x = rxbar, linetype = spec)) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3,
                linetype = "dotted") +
    geom_line(aes(y = bmin), linewidth = 0.55, na.rm = TRUE) +
    geom_line(aes(y = bmax), linewidth = 0.55, na.rm = TRUE) +
    geom_rug(data = ticks, aes(x = x), inherit.aes = FALSE, sides = "b",
              length = grid::unit(0.03, "npc"), linewidth = 0.4) +
    scale_linetype_manual(values = c("rybar = Inf"   = "solid",
                                     "rybar = rxbar" = "dashed")) +
    coord_cartesian(xlim = c(0, 1.4), ylim = c(-2, 6)) +
    labs(x = expression(bar(r)[x]), y = expression(beta[long]),
         linetype = NULL) +
    theme_regsen(base_size = 10)
```

The two zero crossings are the two breakdown points of Table 1, Panel C:

```{r fig1-left-crossings}
c(rybar_inf   = abs(solid$breakdown),     # paper: 0.804
  rybar_rxbar = abs(dashed$breakdown))    # paper: 0.96
```

### Right panel: the breakdown frontier in $(\bar r_X, \bar r_Y)$ space

The paper's right panel is a different object from the left: it plots the
breakdown frontier $\bar r_Y^{bf}(\bar r_X)$, one curve per $\bar c$, in
the $(\bar r_X, \bar r_Y)$ plane. `regsen_breakdown()` solves for the
$\bar r_X$ breakdown at a *fixed* $\bar r_Y$, so sweeping $\bar r_Y$
traces the same curve from the other side:

```{r fig1-right, fig.height = 4.5}
frontier <- function(cbar, rys) {
    rx <- vapply(rys, function(ry) {
        r <- try(regsen_breakdown(form, bfg2020, compare = w1,
                                   cbar = cbar, rybar = ry), silent = TRUE)
        if (inherits(r, "try-error")) NA_real_ else abs(r$results$breakdown[1])
    }, numeric(1))
    data.frame(rxbar = rx, rybar = rys, cbar = factor(cbar))
}

rys <- c(seq(0.6, 2, by = 0.2), 2.5, 3, 4)
fr  <- do.call(rbind, lapply(c(1, 0.75, 0.5), frontier, rys = rys))

ggplot(fr, aes(x = rxbar, y = rybar, group = cbar, colour = cbar)) +
    geom_line(linewidth = 0.6, na.rm = TRUE) +
    scale_colour_regsen() +
    labs(x = expression(bar(r)[x]), y = expression(bar(r)[y]),
         colour = expression(bar(c))) +
    theme_regsen(base_size = 10)
```

Each curve is the boundary between $(\bar r_X, \bar r_Y)$ pairs under
which the sign conclusion survives and those under which it fails. For
$\bar c = 1$ the frontier flattens onto $\bar r_X = 0.804$, the breakdown
point from Table 1; smaller $\bar c$ pushes the frontier out, because
restricting how endogenous the controls may be makes the conclusion
harder to overturn.

**One difference from the published figure.** The paper's right panel runs
to $\bar r_X = 4$. This reconstruction stops near $\bar r_X = 1/\bar c$,
because Lemma S9 of the paper gives $\bar z_X = +\infty$ once
$\bar r_X \bar c \geq 1$, and the package's breakdown search is bounded by
that same quantity. Reaching further along the horizontal arm would mean
solving for $\bar r_Y^{bf}$ directly at a given $\bar r_X$ instead of
inverting, which the package does not currently expose.

### Overlaying the calibration values

```{r fig1-with-rho, fig.height = 5}
df_rho <- calibrate_rho(form, bfg2020, compare = w1)
df_rho$variable <- labels[df_rho$variable]
df_rho$rho_dec  <- df_rho$rho / 100  # express as fraction

# Several covariates calibrate to similar rho, so labels placed at a
# common x collide into an unreadable stack. Staggering them across three
# x positions keeps every label rather than dropping the crowded ones.
df_rho <- df_rho[order(-df_rho$rho_dec), ]
df_rho$xpos <- rep(c(0.02, 0.36, 0.70), length.out = nrow(df_rho))

plot(fig1R) +
    geom_hline(yintercept = df_rho$rho_dec, linetype = "dotted",
                colour = "grey55", linewidth = 0.3) +
    geom_text(data = df_rho,
               aes(x = xpos, y = rho_dec, label = variable),
               hjust = 0, vjust = -0.4, size = 2.5, colour = "grey25") +
    coord_cartesian(ylim = c(0, 1.45))
```

Each dotted line is one covariate's $\rho_k$: how much of the treatment's
variation that single observed control explains. Reading the frontier
against them is the point of the figure -- a covariate sitting *above*
the frontier would, if an unobservable were comparably important, be
enough to overturn the sign.

## Figures 2 and 3

Figure 2 of DMP (2026) reports the same analysis for the *Cut Spending
on Poor* outcome, and Figure 3 reports a sensitivity analysis on the
choice of calibration covariates. Both require additional outcome
variables that are **not** part of the bundled `bfg2020` slice. Once the
full Bazzi-Fiszbein-Gebresilasse replication data is loaded, the calls
look like:

```r
# Hypothetical, given a column `cut_spending_poor`:
regsen_bounds(cut_spending_poor ~ tye_tfe890_500kNI_100_l6 + <controls>,
              data, compare = w1, cbar = 1)
regsen_breakdown(cut_spending_poor ~ ...,
                  data, compare = w1, cbar = seq(0, 1, 0.02))
```

For Figure 3, vary the `compare` argument to include / exclude
calibration covariates:

```r
for (drop in w1) {
    res <- regsen_bounds(form, bfg2020, compare = setdiff(w1, drop),
                          cbar = 1)
    # collect & plot
}
```

## Bonus: identified set under $\bar r_Y < \infty$

The paper notes that restricting the unobservable's effect on the outcome
shrinks the identified set considerably. Reproducing Figure 4 of the
appendix style:

```{r ry-finite}
fy <- regsen_bounds(form, bfg2020, compare = w1, rybar = 2,
                     rxbar = seq(0, 0.95, length.out = 30))
plot(fy, ylim = c(-1, 5),
     title = "rybar = 2 finite-impact constraint")
```
