---
title: "regsensitivity: a tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{regsensitivity: a tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
```

This vignette replicates the BFG2020 empirical application used in the Stata
`regsensitivity` vignette, walking through the **Diegert-Masten-Poirier
(2026)** and **Oster (2019) / Masten-Poirier (2026)** analyses.

```{r}
library(regsensitivity)
library(ggplot2)
data(bfg2020)
bfg2020$statea <- factor(bfg2020$statea)

controls <- c("log_area_2010", "lat", "lon", "temp_mean", "rain_mean",
              "elev_mean", "d_coa", "d_riv", "d_lak", "ave_gyi", "statea")
form <- avgrep2000to2016 ~ tye_tfe890_500kNI_100_l6 +
    log_area_2010 + lat + lon + temp_mean + rain_mean + elev_mean +
    d_coa + d_riv + d_lak + ave_gyi + statea
compare <- setdiff(controls, "statea")
```

The "short" regression coefficient (with no controls) and the "medium"
regression coefficient (with state FE and the comparison controls) give a
first sense of how much controls move the estimate:

```{r}
short <- lm(avgrep2000to2016 ~ tye_tfe890_500kNI_100_l6, data = bfg2020)
med   <- lm(form, data = bfg2020)
coef(short)[2]
coef(med)["tye_tfe890_500kNI_100_l6"]
```

## DMP (2026) bounds

`regsen_bounds()` is the workhorse. Out of the box it sweeps over a grid of
`rxbar` from 0 to the threshold where the identified set first becomes
$(-\infty,+\infty)$, holding `cbar` fixed at the supplied value.

```{r}
bnds <- regsen_bounds(form, bfg2020, compare = compare, cbar = 0.1)
print(bnds)
```

The bounds widen monotonically as `rxbar` (the strength of the unobservable's
effect on the treatment) grows. The breakdown point reported in the header
is the `rxbar` value at which the hypothesis ($\beta>0$, in this case)
first fails.

```{r}
plot(bnds)
```

Sweeping over multiple `cbar` values lets you see how stronger correlations
between the comparison controls and the unobservable affect the bounds:

```{r}
multi <- regsen_bounds(form, bfg2020, compare = compare,
                        rxbar = seq(0, 2, length.out = 21),
                        cbar = c(0, 0.2, 0.4, 0.6, 0.8, 1.0))
plot(multi)
```

### Constrained `rybar`

Setting `rybar` to a finite value adds a constraint on the unobservable's
effect on the outcome. When `cbar > 0` this is a nonconvex problem and we
solve it with a global optimizer (`nloptr` / DIRECT-L); when `cbar = 0` it
has a closed form. Both cases run through the same `regsen_bounds()` call:

```{r}
fin <- regsen_bounds(form, bfg2020, compare = compare, rybar = 2)
print(fin)
```

You can also express `rybar` as a function of `rxbar`, e.g. `rybar = rxbar`:

```{r}
expr <- regsen_bounds(form, bfg2020, compare = compare,
                       rybar_expr = function(rx) rx,
                       rxbar = seq(0, 1, 0.1))
expr$results
```

## DMP breakdown

`regsen_breakdown()` reports the breakdown point as a function of one of
the sensitivity parameters, holding the others fixed:

```{r}
bd <- regsen_breakdown(form, bfg2020, compare = compare,
                        cbar = seq(0, 1, 0.05))
plot(bd)
```

You can also sweep across hypothesis values:

```{r}
hyp <- regsen_breakdown(form, bfg2020, compare = compare,
                         beta = bnd_lb(seq(-1, 1, 0.2)))
print(hyp)
```

## Oster (2019) bounds

Switching to the Oster analysis with `analysis = "oster"`:

```{r}
o <- regsen_bounds(form, bfg2020, compare = compare,
                    analysis = "oster")
print(o)
```

Plotting the equality identified set (rotated so that delta is on the
x-axis and beta on the y-axis):

```{r}
o2 <- regsen_bounds(form, bfg2020, compare = compare,
                     analysis = "oster",
                     delta = seq(-3, 3, 0.05))
plot(o2, ylim = c(-5, 8))
```

### Oster, |delta| <= d (bound)

```{r}
ob <- regsen_bounds(form, bfg2020, compare = compare,
                     analysis = "oster",
                     delta = seq(0, 0.999, 0.001),
                     delta_type = "bound")
plot(ob, ylim = c(-50, 50))
```

## Oster breakdown

```{r}
obd_eq <- regsen_breakdown(form, bfg2020, compare = compare,
                            analysis = "oster",
                            r2long = seq(0.15, 1, 0.05),
                            beta = bnd_eq(0))
plot(obd_eq)

obd_sign <- regsen_breakdown(form, bfg2020, compare = compare,
                              analysis = "oster",
                              r2long = seq(0.15, 1, 0.05))
plot(obd_sign)
```

## Summary call

When you don't know where to start, `regsen_summary()` runs the default
DMP bounds analysis plus an Oster breakdown sweep at a few standard
`r2long` values, matching the behaviour of Stata's `regsensitivity` with
no subcommand.

```{r}
s <- regsen_summary(form, bfg2020, compare = compare)
print(s)
```

## References

* Diegert, P., Masten, M., and Poirier, A. (2026). Assessing Omitted
  Variable Bias when the Controls are Endogenous. arXiv:2206.02303.
* Oster, E. (2019). Unobservable Selection and Coefficient Stability:
  Theory and Evidence. *JBES* 37(2), 187--204.
* Masten, M., and Poirier, A. (2026). The Effect of Omitted Variables on
  the Sign of Regression Coefficients. arXiv:2208.00552.
* Bazzi, S., Fiszbein, M., and Gebresilasse, M. (2020). Frontier
  Culture: The Roots and Persistence of "Rugged Individualism" in the
  United States. *Econometrica* 88(6), 2329--2368.
